---
title: 'STAT 423 Winter 2026 - Final Project'
author: 'Krish Doshi, Lauren Yee, Kian Sharifzadeh'
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document:
    number_sections: true
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MASS)
```


## Data Processing - Krish Doshi
```{r data-cleanup}
orig_data <- read.csv("billboard.csv")

cleaned_data <- data.frame(artist = orig_data$artist, rating = orig_data$overall_rating, 
  age=orig_data$front_person_age)

cleaned_data$gender <- ifelse(orig_data$artist_male == 1, "Male", "Not Male")

cleaned_data$race <- ifelse(orig_data$artist_white == 1, "White", "Not White")

cleaned_data$genre <- ifelse(is.na(orig_data$cdr_genre), "Genre Not Specified", 
    orig_data$cdr_genre)

cleaned_data$genre <- ifelse(grepl("Pop", orig_data$cdr_genre), "Pop", 
                      ifelse(grepl("Country", orig_data$cdr_genre), "Country",
                      ifelse(grepl("Hip-Hop", orig_data$cdr_genre), "Hip-Hop",
                      ifelse(grepl("Rock", orig_data$cdr_genre), "Rock",
                      ifelse(grepl("Jazz", orig_data$cdr_genre), "Jazz", "Other")))))

cleaned_data <- cbind(cleaned_data, bpm=orig_data$bpm, energy=orig_data$energy, 
                danceability = orig_data$danceability, happiness = orig_data$happiness, 
                loudness_db = orig_data$loudness_d_b, 
                length_sec = orig_data$length_sec)


cleaned_data$year <- as.numeric(substr(orig_data$date, 1, 
                    unlist(gregexpr("-", orig_data$date))[1] - 1))

cleaned_data <- cleaned_data[rowSums(is.na(cleaned_data)) == 0 , ]

```

```{r results='hide'}
model_results <- list()


for(i in seq(1960, 2030, 10)){
  decade_label <- paste0(i - 10, "s")

  subpop <- subpop_60s <- cleaned_data[cleaned_data$year >= i - 10 & 
                          cleaned_data$year < i, ]
  
  # --- SAFETY CHECK 1: Ensure enough rows exist ---
  # If you have < 50 rows, the interaction model will likely fail.
  if(nrow(subpop) < 50) {
    message(paste("Skipping", decade_label, "- Not enough data points."))
    next
  }
  
  factor_counts <- sapply(subpop[, c("genre", "gender", "race")], 
                  function(x) length(unique(na.omit(x))))
  
  if(any(factor_counts < 2)) {
    message(paste("Skipping", decade_label, "due to single-level factor:", 
                  names(factor_counts)[factor_counts < 2]))
    next
  }
  
  first_model <- lm(data = subpop, rating ~ age + race + gender + genre + energy 
                  + danceability + happiness + bpm + loudness_db + length_sec)

  second_model <- lm(data = subpop, rating ~ age + gender + race + age*genre 
                + age*energy + age*danceability + age*happiness + age*bpm 
                + age*loudness_db + age*length_sec + race*genre + race*energy 
                + race*danceability + race*happiness + race*bpm + race*loudness_db 
                + race*length_sec + gender*genre + gender*energy + gender*danceability 
                + gender*happiness + gender*bpm +gender*loudness_db + gender*length_sec )

  # --- Plotting (Wrapped in 'try' so one bad plot doesn't stop the whole loop) ---
  # try({
  #   plot(first_model, which = 1:2)
  #   boxcox(first_model, plotit=TRUE)
  #   plot(second_model, which = 1:2)
  #   boxcox(second_model, plotit=TRUE)
  # })
  optimal_model <- step(second_model, direction="backward")
  
  model_results[[decade_label]] <- list(
    "first"   = first_model,
    "second"  = second_model,
    "optimal" = optimal_model
  )
  
  print(decade_label)
  
  summary(first_model)
  
  summary(second_model)
  
  summary(optimal_model)
  
}
```

```{r}
# Loop through each decade in the model_results list
for (decade in names(model_results)) {
  
  # Check if the 'optimal' model exists for this decade
  if (!is.null(model_results[[decade]]$optimal)) {
    
    cat("\n", "========================================", "\n")
    cat("  SUMMARY FOR OPTIMAL MODEL:", decade, "\n")
    cat("========================================", "\n")
    
    # Print the summary
    print(summary(model_results[[decade]]$optimal))
    
  } else {
    cat("\nNo optimal model found for:", decade, "\n")
  }
}
```

## Preliminary Exploration of Data through Visual Analysis

### Plots of Individual Relationships

The plots show the relationship between the predictors and the response variable, rating. 

```{r}
library(ggplot2)
library(patchwork)
plot_age <- ggplot(cleaned_data, aes(x = age, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs Age")

plot_energy <- ggplot(cleaned_data, aes(x = energy, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs Energy")

plot_danceability <- ggplot(cleaned_data, aes(x = danceability, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs Danceability")

plot_happiness <- ggplot(cleaned_data, aes(x = happiness, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs Happiness")

plot_age + plot_energy + plot_danceability + plot_happiness

plot_bpm <- ggplot(cleaned_data, aes(x = bpm, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs BPM")

plot_loudness_db <- ggplot(cleaned_data, aes(x = loudness_db, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs Loudness(db)")

plot_length_sec <- ggplot(cleaned_data, aes(x = length_sec, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs Song Length(sec)")

plot_bpm + plot_loudness_db + plot_length_sec

plot_race <- ggplot(cleaned_data, aes(x = race, y = rating, fill = race)) +
  geom_boxplot() +
  theme_classic() +
  ggtitle("Rating vs Race") + 
  guides(fill = "none")

plot_gender <- ggplot(cleaned_data, aes(x = gender, y = rating, fill = gender)) +
  geom_boxplot() +
  theme_classic() +
  ggtitle("Rating vs Gender") + 
  guides(fill = "none")

plot_genre <- ggplot(cleaned_data, aes(x = genre, y = rating, fill = genre)) +
  geom_boxplot() +
  theme_classic() +
  ggtitle("Rating vs Genre")

plot_race + plot_gender + plot_genre 
```
As expected, banding is observed in the scatterplots because the response variable, rating, is discrete. Since race, gender, and genre are categorical predictors, the relationship with rating is displayed by boxplots instead.

```{r}
ggplot(cleaned_data, aes(x = rating)) +
  geom_histogram(
    binwidth = 0.5,
    boundary = 0.5, 
    fill = "black",  
    color = "grey",
    linewidth = 0.3
  ) +
  labs(
    title = "Histogram of Song Ratings",
    x = "Rating",
    y = "Frequency"
  ) +
  theme_classic()
```
However, the histogram of ratings appears approximately symmetric and unimodal, resembling a normal distribution. While the ratings are discrete (integer-valued) and there are many categorical predictors, the distribution spans multiple levels, creating a shape that approximates continuity. This supports the treating of rating as a continuous response variable in the regression analysis.

### Plots of Grouped Relationships

Data visualizations for grouped relationships help us examine whether relationships between predictors and ratings differ across categories. These plot justify the inclusion of categorical predictors in the regression models and help identify potential interaction effects.

```{r}
model_60s <- model_results[["1960s"]]$optimal
model_70s <- model_results[["1970s"]]$optimal
model_80s <- model_results[["1980s"]]$optimal
model_90s <- model_results[["1990s"]]$optimal
model_2000s <- model_results[["2000s"]]$optimal
model_2010s <- model_results[["2010s"]]$optimal
```

```{r}
vars_by_decade <- lapply(model_results, function(x) {
  attr(terms(x$optimal), "term.labels")
})

all_vars <- unique(unlist(vars_by_decade))

presence_matrix <- sapply(vars_by_decade, function(vars) {
  all_vars %in% vars
})

rownames(presence_matrix) <- all_vars
presence_matrix

rownames(presence_matrix)[rowSums(presence_matrix) == ncol(presence_matrix)]
```
```{r}
ggplot(cleaned_data, aes(x = length_sec, y = rating)) +
  geom_smooth(aes(color = genre), method = "lm", se = FALSE) +
  theme_minimal()
```
The fitted regression lines indicate that the relationship between song length and rating varies by genre. In particular, certain genres (e.g., Other and Country) exhibit steeper positive slopes, while others (e.g., Rock and Pop) show weaker or slightly negative trends, suggesting a potential interaction between length and genre.

```{r}
model_60s
model_70s 
model_80s 
model_90s 
model_2000s 
model_2010s

anova(model_results[["1960s"]]$first, model_60s)
anova(model_60s, model_results[["1960s"]]$second)

anova(model_results[["1970s"]]$first, model_70s)
anova(model_70s, model_results[["1970s"]]$second)

anova(model_results[["1980s"]]$first, model_80s)
anova(model_80s, model_results[["1980s"]]$second)

anova(model_results[["1990s"]]$first, model_90s)
anova(model_90s, model_results[["1990s"]]$second)

anova(model_results[["2000s"]]$first, model_2000s)
anova(model_2000s, model_results[["2000s"]]$second)

anova(model_results[["2010s"]]$first, model_2010s)
anova(model_2010s, model_results[["2010s"]]$second)
```

```{r}
    #remove eval = FALSE option after filling in the blanks.
    
    ggplot(data = chickwts_ci)+
      geom_segment( mapping = aes(x = lower, 
                                  xend = upper,
                                  y = feed,
                                  yend = feed) ) +
      labs(title = "95% Confidence Interval of Chicken Weight (g) vs Feed",
               x = "Chicken Weight (g)") +
      scale_x_continuous(breaks = seq(125, 375, 25))

```

