---
title: 'STAT 423 Winter 2026 - Final Project'
author: 'Krish Doshi, Lauren Yee, Kian Sharifzadeh'
date: '`r Sys.Date()`'
output:
  pdf_document: default
  html_document:
    number_sections: true
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MASS)
```


## Data Processing - Krish Doshi
```{r data-cleanup}
orig_data <- read.csv("billboard.csv")

cleaned_data <- data.frame(artist = orig_data$artist, rating = orig_data$overall_rating, age=orig_data$front_person_age)

cleaned_data$gender <- ifelse(orig_data$artist_male == 1, "Male", "Not Male")

cleaned_data$race <- ifelse(orig_data$artist_white == 1, "White", "Not White")

cleaned_data$genre <- ifelse(is.na(orig_data$cdr_genre), "Genre Not Specified", orig_data$cdr_genre)

cleaned_data$genre <- ifelse(grepl("Pop", orig_data$cdr_genre), "Pop", 
                             ifelse(grepl("Country", orig_data$cdr_genre), "Country",
                                    ifelse(grepl("Hip-Hop", orig_data$cdr_genre), "Hip-Hop",
                                           ifelse(grepl("Rock", orig_data$cdr_genre), "Rock",
                                                  ifelse(grepl("Jazz", orig_data$cdr_genre), "Jazz", "Other")))))

cleaned_data <- cbind(cleaned_data, bpm=orig_data$bpm, energy=orig_data$energy, danceability = orig_data$danceability, happiness = orig_data$happiness, loudness_db = orig_data$loudness_d_b, length_sec = orig_data$length_sec)


#Wind
cleaned_data$has_winds <- ifelse(orig_data$clarinet, "True",
                                 ifelse(orig_data$flute_piccolo, "True",
                                        ifelse(orig_data$harmonica, "True",
                                               ifelse(orig_data$kazoo, "True",
                                                      ifelse(orig_data$ocarina, "True",
                                                             ifelse(orig_data$saxophone, "True", 
                                                            ifelse(orig_data$accordion, "True", "False")))))))


#Brass
cleaned_data$has_brass <- ifelse(orig_data$trumpet, "True", "False")

#Percussion
cleaned_data$has_percussion <- ifelse(orig_data$piano_keyboard_based, "True",
                                 ifelse(orig_data$bongos, "True",
                                        ifelse(orig_data$cowbell, "True",
                                               ifelse(orig_data$handclaps_snaps, "True", "False"))))

#Vocal or Whistle
cleaned_data$has_vocal_or_whistle <- ifelse(orig_data$vocally_based, "True",
                                 ifelse(orig_data$falsetto_vocal, "True",
                                        ifelse(orig_data$human_whistling, "True", "False")))

#String
cleaned_data$has_strings <- ifelse(orig_data$bass_based, "True",
                                 ifelse(orig_data$guitar_based, "True",
                                        ifelse(orig_data$orchestral_strings, "True",
                                               ifelse(orig_data$banjo, "True",
                                                      ifelse(orig_data$mandolin, "True",
                                                             ifelse(orig_data$pedal_lap_steel, "True",
                                                              ifelse(orig_data$sitar, "True",
                                                              ifelse(orig_data$ukulele, "True",
                                                              ifelse(orig_data$violin, "True","False")))))))))


cleaned_data$year <- as.numeric(substr(orig_data$date, 1, unlist(gregexpr("-", orig_data$date))[1] - 1))

cleaned_data <- cleaned_data[rowSums(is.na(cleaned_data)) == 0 , ]

cutoff <- round(min(cleaned_data$year) + ((max(cleaned_data$year) - min(cleaned_data$year)) / 2))

first_subpop <- cleaned_data[cleaned_data$year < cutoff, ]

second_subpop <- cleaned_data[cleaned_data$year >= cutoff, ]




```

## Models - Kian Sharifzadeh
```{r}
model_subpop1_1 <- lm(data = first_subpop, rating ~ age + race + gender + genre + energy + danceability + happiness + bpm + loudness_db + length_sec)
summary(model_subpop1_1)
```

```{r}
#residual anaylsis plots for subpop 1
plot(model_subpop1_1, which = 1)
plot(model_subpop1_1, which = 2)
boxcox(model_subpop1_1, plotit=T)
```

```{r}
step(model_subpop1_1, direction="backward")
```

```{r}
new_model_1 <- lm(data = cleaned_data, log(rating) ~ age+ gender+ race + log(bpm) + genre*log(energy) + genre*log(danceability) + happiness +length_sec + loudness_db + has_brass + has_percussion  + has_vocal_or_whistle + year*log(energy) + age*genre)
summary(new_model_1)

```



```{r}
model_subpop1_2 <- lm(data = first_subpop, rating ~ age + gender + race + age*genre + age*energy + age*danceability + age*happiness + age*bpm + age*loudness_db + age*length_sec + race*genre + race*energy + race*danceability + race*happiness + race*bpm + race*loudness_db + race*length_sec + + gender*genre + gender*energy + gender*danceability + gender*happiness + gender*bpm +gender*loudness_db + gender*length_sec )
summary(model_subpop1_2)

```

```{r}
#residual anaylsis plots for subpop 2
plot(model_subpop1_2, which = 1)
plot(model_subpop1_2, which = 2)
boxcox(model_subpop1_2, plotit=T)

```

```{r}
step(model_subpop1_2, direction="backward")
```

```{r}
new_model_subpop1_2 <- lm(data = first_subpop, rating ~ age + gender + race + genre + energy + danceability+ happiness + loudness_db + length_sec  + age*energy + age*danceability  + race*genre + race*length_sec + gender*genre + gender*energy + gender*danceability + gender*happiness  + gender*length_sec )
summary(new_model_subpop1_2)

```

```{r}
model_subpop2_1 <- lm(data = second_subpop, rating ~ age + race + gender + genre + energy + danceability + happiness + bpm + loudness_db + length_sec)
summary(model_subpop2_1)
```

```{r}
plot(model_subpop2_1, which = 1)
plot(model_subpop2_1, which = 2)
boxcox(model_subpop2_1, plotit=T)
```

```{r}
model_subpop2_2 <- lm(data = second_subpop, rating ~ age + gender + race +age*genre + age*energy + age*danceability + age*happiness + age*bpm + age*loudness_db + age*length_sec + race*genre + race*energy + race*danceability + race*happiness + race*bpm + race*loudness_db + race*length_sec + + gender*genre + gender*energy + gender*danceability + gender*happiness + gender*bpm +gender*loudness_db + gender*length_sec)
summary(model_subpop2_2)

```

```{r}
step(model_subpop2_2, direction="backward")

```

```{r}
plot(model_subpop2_2, which = 1)
plot(model_subpop2_2, which = 2)
boxcox(model_subpop2_2, plotit=T)
```

```{r}
model_total_1 <- lm(data = cleaned_data, rating ~ age + race + gender + genre + energy + danceability + happiness + bpm + loudness_db + length_sec)
summary(model_total_1)

```

```{r}
plot(model_total_1, which = 1)
plot(model_total_1, which = 2)
boxcox(model_total_1, plotit=T)

```

```{r}
model_total_2 <- lm(data = cleaned_data, rating ~ age + gender + race +age*genre + age*energy + age*danceability + age*happiness + age*bpm + age*loudness_db + age*length_sec + race*genre + race*energy + race*danceability + race*happiness + race*bpm + race*loudness_db + race*length_sec + + gender*genre + gender*energy + gender*danceability + gender*happiness + gender*bpm +gender*loudness_db + gender*length_sec )
summary(model_total_2)

```

```{r}


step(model_total_2, direction="backward")

```


```{r}
anova(model_subpop1_1, model_subpop1_2)

```

```{r}
anova(model_subpop2_1, model_subpop2_2)

```


```{r}
anova(model_total_1, model_total_2)

```

```{r}
plot(model_total_2, which = 1)
plot(model_total_2, which = 2)
boxcox(model_total_2, plotit=T)
```

## Preliminary exploration of data through visual analysis

```{r}
library(ggplot2)

ggplot(cleaned_data, aes(x = rating)) +
  geom_histogram(
    binwidth = 0.5,
    boundary = 0.5, 
    fill = "blue",  
    color = "black",
    linewidth = 0.3
  ) +
  labs(
    title = "Histogram of Song Ratings",
    x = "Rating",
    y = "Frequency"
  ) +
  theme_classic()
```
The histogram of ratings appears approximately symmetric and unimodal, resembling a normal distribution. Although the ratings are discrete (integer-valued), the distribution spans multiple levels, creating a shape that approximates continuity. This supports the treating of rating as a continuous response variable in the regression analysis.

```{r}
library(patchwork)
p1 <- ggplot(first_subpop, aes(x = age, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs Age")

p2 <- ggplot(first_subpop, aes(x = energy, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs Energy")

p3 <- ggplot(first_subpop, aes(x = danceability, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs Danceability")

p1 + p2 + p3

p4 <- ggplot(first_subpop, aes(x = happiness, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs Happiness")

p5 <- ggplot(first_subpop, aes(x = bpm, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs BPM")

p6 <- ggplot(first_subpop, aes(x = loudness_db, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs Loudness")

p4 + p5 + p6

p7 <- ggplot(first_subpop, aes(x = length_sec, y = rating)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = F) +
  theme_classic() +
  ggtitle("Rating vs Song Length")

p7
```